MSSM-Penn-Pitt - DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Model Selection

### 2020-05-14 individual-cohorts/MSSM-Penn-Pitt_DLPFC_FEATURECOUNTS_GRCh38_CQN.Rmd
test DLPFC cohort

---
title: "MSSM-Penn-Pitt - ACC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Model Selection"
author: "Kelsey Montgomery, Gabriel Hoffman, Thanneer Perumal"
date: "`r date()`"
output: html_document
---

@@

knit2synapse::createAndKnitToFolderEntity(file = "individual-cohorts/MSSM-Penn-Pitt_ACC_FEATURECOUNTS_GRCh38_CQN.Rmd",
                                 parentId ="syn22012942",
                                 folderName = "MSSM-Penn-Pitt - ACC - DREAM - FEATURECOUNTS - GRCh38 - CQN")
```

@@

```{r synapse.parameters, include=FALSE, cache=FALSE}
parentId = 'syn22012942';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of GRCh38 feature counts with CQN normalisation (ACC)';

thisFileName <- 'MSSM-Penn-Pitt_ACC_FEATURECOUNTS_GRCh38_CQN.Rmd'

@@

METADATA = metadata %>%
  dplyr::left_join(GENOTYPE) %>%
  dplyr::rename(PMI = `PMI_(in_hours)`,
                ReportExclude = `rnaSeq_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                Tissue = `rnaSeq_dissection:Brain_Region`,
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, 
                IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, 
                GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, 
                rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, 
                AlignmentRate = `rnaSeq_report:Percent_Aligned`,
                RIN = `rnaSeq_isolation:RIN`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, 
                RIN, ReportExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, RibozeroBatch, Tissue, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate) %>% 
  dplyr::filter(Institution %in% c("MSSM", "Penn", "Pitt")) %>% 
  dplyr::filter(Tissue %in% c("ACC")) %>% 
  dplyr::filter(SampleID %in% colnames(counts), !is.na(SampleID)) %>%
  dplyr::mutate(Dx = forcats::fct_recode(Dx,  Other = "AFF", Other = "BP", Control = "Control", SCZ = "SCZ"))
```

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, "SampleID")) %>%
  dplyr::mutate(label = SampleID) %>% 
  dplyr::mutate(label = ifelse((label %in% outliers), label, NA)) %>% 
  dplyr::mutate(Tissue = "ACC")

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>% 
  dplyr::mutate(Tissue = "ACC")

@@

# Stepwise regression
phase1 <- mvIC::mvForwardStepwise(exprObj = CQN.GENE_EXPRESSION$E,
                         baseFormula = base_formula,
                         data = COVARIATES,
                         variables = array(c("(1|Institution)", "(1|Reported_Gender)", "scale(PMI)",
                                                             "scale(ageOfDeath)", "scale(RIN)", "scale(RIN2)", "scale(EV.1)",
                                                             "scale(EV.2)", "scale(EV.3)", "scale(EV.4)", "scale(EV.5)")))
kable(phase1)
```

#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
phase2 <- mvIC::mvForwardStepwise(exprObj = CQN.GENE_EXPRESSION$E,
                         baseFormula = phase1$formula, 
                         data = COVARIATES,
                         variables = array(c("(1|LibraryBatch)", "(1|FlowcellBatch)")))
kable(phase2)
```

#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
phase3 <- mvIC::mvForwardStepwise(exprObj = CQN.GENE_EXPRESSION$E,
                         baseFormula = phase2$formula, 
                         data = COVARIATES,
                         variables = array(c("scale(IntragenicRate)", "scale(IntronicRate)","scale(IntergenicRate)","scale(rRNARate)", 
                                             "scale(TotalReads)", "scale(GenesDetected)", "scale(MappedReads)")))
kable(phase3)
```

@@

The model for clinical, technical and batch variables selected after phase I, II and III is `r phase3$formula`

@@

# Code
CODE <- Folder(name = "MSSM-Penn-Pitt - ACC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'CMC_ACC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = synapser::File('CMC_ACC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(COV_OBJ, annotations = all.annotations)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(COUNT_OBJ, annotations = all.annotations)

# Store logCPM
CQN.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('CMC_ACC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(LCOUNT_OBJ, annotations = all.annotations)

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('CMC_ACC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
OFFSET_OBJ = synStore(OFFSET_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(OFFSET_OBJ, annotations = all.annotations)
```

>>>>

---
title: "MSSM-Penn-Pitt - DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Model Selection"
author: "Kelsey Montgomery, Gabriel Hoffman, Thanneer Perumal"
date: "`r date()`"
output: html_document
---

@@

knit2synapse::createAndKnitToFolderEntity(file = "individual-cohorts/MSSM-Penn-Pitt_DLPFC_FEATURECOUNTS_GRCh38_CQN.Rmd",
                                 parentId ="syn22012942",
                                 folderName = "MSSM-Penn-Pitt - DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN")
```

@@

```{r synapse.parameters, include=FALSE, cache=FALSE}
parentId = 'syn22012942';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of GRCh38 feature counts with CQN normalisation (DLPFC)';

thisFileName <- 'MSSM-Penn-Pitt_DLPFC_FEATURECOUNTS_GRCh38_CQN.Rmd'

@@

METADATA = metadata %>%
  dplyr::left_join(GENOTYPE) %>%
  dplyr::rename(PMI = `PMI_(in_hours)`,
                ReportExclude = `rnaSeq_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                Tissue = `rnaSeq_dissection:Brain_Region`,
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, 
                IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, 
                GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, 
                rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, 
                AlignmentRate = `rnaSeq_report:Percent_Aligned`,
                RIN = `rnaSeq_isolation:RIN`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, 
                RIN, ReportExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, RibozeroBatch, Tissue, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate) %>% 
  dplyr::filter(Institution %in% c("MSSM", "Penn", "Pitt")) %>% 
  dplyr::filter(Tissue %in% c("DLPFC")) %>% 
  dplyr::filter(SampleID %in% colnames(counts), !is.na(SampleID)) %>%
  dplyr::mutate(Dx = forcats::fct_recode(Dx,  Other = "AFF", Other = "BP", Control = "Control", SCZ = "SCZ"))
```

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, "SampleID")) %>%
  dplyr::mutate(label = SampleID) %>% 
  dplyr::mutate(label = ifelse((label %in% outliers), label, NA)) %>% 
  dplyr::mutate(Tissue = "DLPFC")

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>% 
  dplyr::mutate(Tissue = "DLPFC")

@@

# Stepwise regression
phase1 <- mvIC::mvForwardStepwise(exprObj = CQN.GENE_EXPRESSION$E,
                         baseFormula = base_formula,
                         data = COVARIATES,
                         variables = array(c("(1|Institution)", "(1|Reported_Gender)", "scale(PMI)",
                                                             "scale(ageOfDeath)", "scale(RIN)", "scale(RIN2)", "scale(EV.1)",
                                                             "scale(EV.2)", "scale(EV.3)", "scale(EV.4)", "scale(EV.5)")))
kable(phase1$settings)
kable(phase1$trace)
as.character(phase1$formula)[2]
```

#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
phase2 <- mvIC::mvForwardStepwise(exprObj = CQN.GENE_EXPRESSION$E,
                         baseFormula = phase1$formula, 
                         data = COVARIATES,
                         variables = array(c("(1|LibraryBatch)", "(1|FlowcellBatch)")))
kable(phase2$settings)
kable(phase2$trace)
as.character(phase2$formula)[2]
```

#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
phase3 <- mvIC::mvForwardStepwise(exprObj = CQN.GENE_EXPRESSION$E,
                         baseFormula = phase2$formula, 
                         data = COVARIATES,
                         variables = array(c("scale(IntragenicRate)", "scale(IntronicRate)","scale(IntergenicRate)","scale(rRNARate)", 
                                             "scale(TotalReads)", "scale(GenesDetected)", "scale(MappedReads)")))
kable(phase3$settings)
kable(phase3$trace)
as.character(phase3$formula)[2]
```

@@

The model for clinical, technical and batch variables selected after phase I, II and III is `r as.character(phase3$formula)[2]`

@@

# Code
CODE <- Folder(name = "MSSM-Penn-Pitt - DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'CMC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = synapser::File('CMC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(COV_OBJ, annotations = all.annotations)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(COUNT_OBJ, annotations = all.annotations)

# Store logCPM
CQN.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('CMC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(LCOUNT_OBJ, annotations = all.annotations)

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('CMC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
OFFSET_OBJ = synStore(OFFSET_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(OFFSET_OBJ, annotations = all.annotations)
```
